---
title: "The Effects of Weather Patterns on MTA Transit Behavior in New York City"
output: html_document
date: "December 6, 2025"
---

Jefferson Remo (jr4550),
Jonathan Huynh (jkh2157),
Nelson Gaillard (ng3005),
Mahitha Jangeti (mj3229),
Matariya Rattanapan (mkr2158)

## Motivation

The COVID-19 pandemic altered patterns of public transit use, exposing the sensitivity of urban mobility to both public-health and environmental conditions. Examining how MTA ridership fluctuated with changing case counts and weather conditions can help identify which factors most strongly affected transit demand, and how long recovery took across time and boroughs. These insights can guide policy around public health preparedness, transit safety, and infrastructure planning under climate variability.

## Related Work

## Initial Questions





## Exploratory Analyses

## Additional Analyses

## Discussion


```{r setup, include=FALSE}
library(lubridate)
library(tidyverse)
library(plotly)
library(patchwork)
library(dplyr)
library(scales)
library(knitr)
library(broom)
knitr::opts_chunk$set(fig.width = 10, fig.height = 6)
```

# Data: Source, Scraping, Cleaning, etc

## Individual Analyses 

## COVID-19 ANALYSIS

COVID-19 data was pulled from opendata.cityofnewyork.us, a site consisting of open data sources about various topics affecting New York City boroughs. 

#### Reading in the covid datafile.
```{r}
covid_df =
read.csv("data/covid_19.csv")
```

#### Separating the date variable into month, day, and year + arranging
```{r}
clean_covid_df =
  covid_df |>
  janitor::clean_names() |>
  mutate(
    date = lubridate::mdy(date_of_interest),
    case_count = as.numeric(gsub(",", "", case_count)),
    case_count_7day_avg = as.numeric(gsub(",", "", case_count_7day_avg)),
    hospitalized_count = as.numeric(gsub(",", "", hospitalized_count)),
    death_count = as.numeric(gsub(",", "", death_count))
  ) |>
  arrange(date)
```

To preemptively ensure consistency across all four data sets, date information was pulled from the data set and mutated to develop a new date variable `date_of_interest`. Additionally, `case_count`, `case_count_7day_avg`, `hospitalized_count`, and `death_count` were all formatted as numeric variables. Lastly, the data set was sorted by date.

#### Marking events for lockdown, rollout of vaccine, and the surge of omicron
```{r}
events = tibble(
  event = c("Lockdown", "Vaccine rollout", "Omicron surge"),
  start = as.Date(c("2020-03-15", "2020-12-14", "2021-12-01")),
  end   = as.Date(c("2020-06-15", "2021-03-01", "2022-02-15"))
)

max_y <- max(clean_covid_df$case_count, na.rm = TRUE)
```

Important events pertaining to the COVID-19 pandemic were highlighted in a new dataframe `events`, marking the estimated start date and estimated end date of `Lockdown`, `Vaccine rollout`, and `Omicron surge`, respectively. 

#### Plotting the daily COVID-19 cases by Year
```{r}
clean_covid_df |> 
  plot_ly(
    x = ~date,
    y = ~case_count,
    type = "scatter",
    mode = "markers+lines",
    marker = list(
      color = ~case_count,
      colorscale = "Viridis",
      size = 5,
      opacity = 0.8,
      colorbar = list(title = "Daily Cases")
    ),
    line = list(color = "gray60", width = 0.5)
  ) |>
  layout(
    title = "NYC Daily COVID-19 Cases (Log Scale)",
    xaxis = list(
      title = "Year",
      range = c(as.Date("2019-01-01"), as.Date("2025-12-31"))
    ),
    yaxis = list(
      title = "Daily COVID-19 Cases (log scale)",
      type = "log"
    ),
    shapes = lapply(1:nrow(events), function(i) {
      list(
        type = "rect",
        xref = "x",
        x0 = events$start[i],
        x1 = events$end[i],
        yref = "paper",
        y0 = 0,
        y1 = 1,
        fillcolor = "rgba(255, 0, 0, 0.15)",
        line = list(width = 0),
        layer = "below"
      )
    }),
    annotations = lapply(1:nrow(events), function(i) {
      list(
        x = events$start[i] + (events$end[i] - events$start[i]) / 2,
        y = 1.03,
        xref = "x",
        yref = "paper",
        text = events$event[i],
        showarrow = FALSE,
        font = list(size = 12)
      )
    })
  )
```

Before diving deeper into further analyses, it is important to visualize COVID-19 cases across the years of interest. In a plotly, `Daily COVID-19 Cases` was plotted against `Year` using the `case_count` and `date` variables. `Daily COVID-19 Cases` was transformed to the log scale to more easily visualize the peaks/spikes of COVID-19 cases from 2020 to 2025. 

Overall, the log-scale plot shows NYC's daily COVID-19 cases from 2020 to 2025, with darker colors representing lower casse counts and bright yellow marking the highest case days. The largest spike appears during the Omicron surge of the pandemic, far exceeding the earlier lockdown and vaccine-rollout waves, even after log transformation. Ultimately, case levels drop substantially after the Omicron surge and settle into smaller, seasonal increases--particularly during Winter--through 2023 to 2025. This insight indicates a sustained decline in large outbreaks.

#### Plot Weekly COVID-19 Cases by Year
```{r}
clean_covid_df |> 
  plot_ly(
    x = ~date,
    y = ~case_count_7day_avg,
    type = "scatter",
    mode = "markers+lines",
    marker = list(
      color = ~case_count_7day_avg,
      colorscale = "Viridis",
      colorbar = list(title = "Weekly Avg Cases"),
      size = 5,
      opacity = 0.6
    ),
    line = list(
      color = "gray50",
      width = 0.6
    ),
    name = "7-day Average"
  ) |>
  layout(
    title = "NYC Weekly COVID-19 Cases (7-day Average, Log Scale)",
    xaxis = list(
      range = c(as.Date("2019-01-01"), as.Date("2025-12-31")),
      tickvals = as.Date(c("2019-01-01","2020-01-01","2021-01-01",
                           "2022-01-01","2023-01-01","2024-01-01","2025-01-01")),
      ticktext = c("2019","2020","2021","2022","2023","2024","2025"),
      title = "Year"
    ),
    yaxis = list(
      title = "7-day Average COVID-19 Cases (log scale)",
      type = "log"
    ),
    shapes = lapply(1:nrow(events), function(i) {
      list(
        type = "rect",
        xref = "x",
        x0 = events$start[i],
        x1 = events$end[i],
        yref = "paper",
        y0 = 0,
        y1 = 1,
        fillcolor = "rgba(255, 0, 0, 0.15)",
        line = list(width = 0),
        layer = "below"
      )
    }),
    annotations = lapply(1:nrow(events), function(i) {
      list(
        x = events$start[i] + (events$end[i] - events$start[i]) / 2,
        y = 1.0,
        xref = "x",
        yref = "paper",
        text = events$event[i],
        showarrow = FALSE,
        font = list(size = 12)
      )
    })
  )
```

COVID-19 over time was also visualized using the `case_count_7day_avg` variable for a more zoomed out picture of COVID-19 from 2020 to 2025. 

After the initial lockdown period, the weekly average shows a sharp and sustained decline in cases, reflecting the immediate impact of reduced mobility and strict public health measures. A similar decline appears following vaccine-rollout in early 2021, where cases fall noticeably, reflecting the early effects of widespread vaccination and reduced transmission. The conclusions from this plot (`7-day Average COVID-19 Cases` vs. `Year`) align with those showcased in the plot above (`Daily COVID-19 Cases` vs. `Year`). 

#### Correlation Heat Map of Cases, Hospitalization, and Deaths
```{r}
corr_matrix =
  clean_covid_df |>
    select(
      Cases = case_count,
      Hospitalizations = hospitalized_count,
      Deaths = death_count
    ) |>
    cor(use = "complete.obs")

plot_ly(
  x = colnames(corr_matrix),
  y = colnames(corr_matrix),
  z = corr_matrix,
  type = "heatmap",
  colorscale = "YlOrRd_r",
  zmin = min(corr_matrix),
  zmax = max(corr_matrix),
  text = round(corr_matrix, 2),
  texttemplate = "%{text}",
  textfont = list(color = "black", size = 16),
  showscale = TRUE
) |>
  layout(
    title = "Correlation Between NYC COVID-19 Cases, Hospitalizations, and Deaths",
    xaxis = list(title = ""),
    yaxis = list(title = "")
  )
```

To visualize associations between COVID-19 cases, hospitalizations, and deaths, a correlation matrix was generated.

The matrix shows that hospitalizations and deaths are strongly correlated, which reflects the expected clinical progression from severe illness to mortality. Cases have a moderate correlation with hospitalizations and a weak corelation with deaths, likely due to widespread vaccination, changes in testing behavior, and milder Omicron waves that reduced the proportion of severe cases.

## WEATHER ANALYSIS

Weather data was pulled from ncei.noaa.gov, a site consisting of open data sources regarding weather patterns across the United States. Since weather is measured from multiple sites throughout NYC, Central Park was used as a proxy for all of NYC.

#### Reading in weather data
```{r}
nyc_weather =
read.csv("data/nyc_weather.csv")
```

#### Cleaning up weather data
```{r}
weather_df =
  nyc_weather |> 
  janitor::clean_names() |> 
  mutate(
    date = as.Date(date),
    year = lubridate::year(date),
    month = lubridate::month(date),
    month = case_match(month,
                       1 ~ "January",
                       2 ~ "February",
                       3 ~ "March", 
                       4 ~ "April",
                       5 ~ "May",
                       6 ~ "June",
                       7 ~ "July",
                       8 ~ "August",
                       9 ~ "September", 
                       10 ~ "October",
                       11 ~ "November",
                       12 ~ "December"
    ),
    tmin_c = tmin / 10,
    tmax_c = tmax / 10,
    average_temp = (tmin_c + tmax_c) / 2,
  ) |> 
  filter(year %in% c(2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025))
```

After weather data was loaded, the `date` variable was coded as a date, and the year and month were extracted such that the month name replaced the month number. This process makes faceting and labeling easier when developing plots. Additionally, originally, the data was stored in the data set as tenths of degrees Celsius (e.g. 250 -> 25°C), so `tmin_c` and `tmax_c` were created to showcase the accurate temperature in degrees Celsius. The average temperature (`average_temp`) was calculated by taking the mean of `tmin_c` and `tmax_c`. Lastly, the data was restricted to include only years of interest: 2018 to 2025.

#### Plot Average Temperature by Year
```{r}
weather_df |> 
  plot_ly(
    x = ~date,
    y = ~average_temp,
    type = "scatter",
    mode = "markers",
    marker = list(
      color = ~average_temp,
      colorscale = "Viridis",
      colorbar = list(title = "Temperature (°C)"),
      size = 5,
      opacity = 0.7
    )
) |>
  layout(
    title = "NYC Average Temperature by Year (2018-2025)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average Temperature (°C)")
  )
```

`Average Temperature (°C)` was plotted from 2018 to 2025. This plot shows a clear and consistent seasonal pattern in NYC temperatures (using Central Park as a proxy), with colder winters (often with temperatures below 0°C) and warm summers that regularly exceed 25-30°C. 

#### Plot Snow Distribution by Year
```{r}
weather_df |>
  mutate(
    month = month(date, label = TRUE, abbr = FALSE),
    month = factor(month, 
                   levels = c("December", "January", "February", "March"))
  ) |>
  filter(
    snow > 0,
    month %in% c("December", "January", "February", "March")
  ) |>
  group_by(year, month) |>
  summarise(total_snow = sum(snow), .groups = "drop") |> 
  ggplot(aes(x = month, y = total_snow, fill = month)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~year, ncol = 2) +
  labs(
    title = "NYC Monthly Snowfall (mm) by Year (2018-2025)",
    x = "Month",
    y = "Total Snowfall (mm)",
    fill = ""
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

`Total Snowfall (mm)` by year was also plotted, however NYC snowfall is charted only for December through March because these four winter months account for virtually all measurable snowfall, thus making the analysis more meaningful.

Across these winters, snowfall is highly variable. 2018 and 2021 stand out as having the snowiest winters. 2019, 2020, 2024, and 2025 show more moderate (but still noticeable) snow accumulation spread across one or two winter months. 2022 has relatively light snowfall, and 2023 is nearly snowfall-free. The discrepancy in snowfall consistency highlights how NYC winters have become increasingly inconsistent. 

#### Plot Rainfall Distribution
```{r}
weather_df |>
  mutate(
    month = month(date, label = TRUE, abbr = FALSE),
    month = factor(month, 
                   levels = c("December", "January", "February", "March")),
    prcp_mm = prcp / 10
  ) |>
  filter(
    month %in% c("December", "January", "February", "March")
  ) |>
  group_by(year, month) |>
  summarise(total_prcp = sum(prcp_mm), .groups = "drop") |> 
  ggplot(aes(x = month, y = total_prcp, fill = month)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~year, ncol = 2) +
  labs(
    title = "NYC Monthly Precipitation (mm) by Year (2018-2025)",
    x = "Month",
    y = "Total Precipitation (mm)",
    fill = ""
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Similarly, `Total Precipitation (mm)` (i.e. rainfall) by year was plotted.

Most years had precipitation concentrated around December and March, while January and February vary more widely. Some winters--such as 2018, 2019, and 2020--depict consistently high precipitation across all four months, whereas other years like 2020 and 2025 appear drier overall.

## MTA ANALYSIS

MTA data was pulled from data.gov, a site consisting of open data from the US government.  Both daily and monthly MTA data was imported.

#### Read in MTA data
```{r}
daily_mta = 
  read.csv("data/mta.csv")

monthly_mta =
  read.csv("data/monthly_mta.csv")
```

#### Clean monthly and daily MTA datasets & join
```{r}
monthly_pre =
  monthly_mta |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(month),
    year = year(date),
    month_num = month(date),
    month_name = month(date, label = TRUE, abbr = FALSE),
    ridership = parse_number(ridership),

    mode = case_when(
      agency %in% c("NYCT Bus", "MTA Bus", "Bus") ~ "Bus",
      agency %in% c("Subway","MNR","LIRR","AAR","SIR") ~ agency,
      TRUE ~ NA_character_
    ),

    days_in_month = days_in_month(date),
    avg_daily_ridership = ridership / days_in_month
  ) |>
  filter(
    year >= 2018,
    year <= 2019,
    !is.na(mode)
  ) |>
  select(year, month_num, month_name, mode, avg_daily_ridership)

daily_mta_clean =
  daily_mta |>
  janitor::clean_names() |>
  mutate(
    date = mdy(date),
    year = year(date),
    month_num = month(date),
    month_name = month(date, label = TRUE, abbr = FALSE),
    count = parse_number(count),

    mode = case_when(
      mode %in% c("NYCT Bus", "MTA Bus", "Bus") ~ "Bus",
      mode %in% c("AAR","Subway","LIRR","MNR","SIR") ~ mode,
      TRUE ~ NA_character_
    )
  ) |>
  filter(year >= 2020, !is.na(mode)) |> 
  mutate(
    weekday = wday(date, label = TRUE),
    weekend = ifelse(weekday %in% c("Sat", "Sun"), "Weekend", "Weekday")
  )

monthly_post =
  daily_mta_clean |>
  group_by(year, month_num, month_name, mode) |>
  summarise(
    avg_daily_ridership = mean(count, na.rm = TRUE),
    .groups = "drop"
  )

full_mta =
  bind_rows(monthly_pre, monthly_post) |>
  arrange(year, month_num, mode) |>
  mutate(
    date = as.Date(paste(year, month_num, "01", sep = "-")),
    period = case_when(
      year < 2020 ~ "Pre-Pandemic",
      year == 2020 & month_num <= 2 ~ "Pre-Pandemic",
      year == 2020 & month_num >= 3 ~ "Pandemic",
      year == 2021 ~ "Pandemic",
      year >= 2022 ~ "Post-Pandemic"),
    period = factor(period, levels = c("Pre-Pandemic", "Pandemic", "Post-Pandemic"))
  )
```

To clean the data, `monthly_mta` was divided into data before the COVID-19 pandemic (`monthly_pre`) and data during/after the COVID-19 pandemic (`monthly_post`). Like for COVID-19 data and Weather data, `year` and `month` were both extracted. Additionally, transit modes were standardized (i.e. combining various bus agencies into `Bus`). An `average_daily_ridership` variable was created by taking monthly ridership data and divided it by the number of days in each respective month. For `monthly_pre`, only observations from years 2018 and 2019 were kept.

The `daily_mta` dataset was cleaned in a similar way by converting dates (extracting year and month), parsing ridership counts, and standardizing transit modes to be consistent with `monthly_pre`. Weekend indicators were also added. Cleaned daily data was then aggregated to `monthly_post` by averaging daily ridership data across each month.

Finally, pre-pandemic and post-2020 monthly datasets are combined in `full_mta`, sorted chronologically, assigned a standardized date (beginning on the first of each month), and organized into three periods--pre-pandemic, pandemic, and post-pandemic--based on year and month. 

#### Pre- vs. Post-Pandemic Modeling
```{r}
pre =
full_mta |> 
  filter(
    period == "Pre-Pandemic"
  ) |> 
  ggplot(aes(x = date, y = avg_daily_ridership)) +
  geom_line(linewidth = 1) +
  facet_wrap(~mode, scales = "free_y") +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  labs(
    title = "Pre-Pandemic Ridership Trends (2018–Feb 2020)",
    x = "Month",
    y = "Average Daily Ridership"
  ) +
  theme_minimal()

pandemic =
full_mta |> 
  filter(
    period == "Pandemic"
  ) |> 
ggplot(aes(x = date, y = avg_daily_ridership)) +
  geom_line(linewidth = 1, color = "red") +
  facet_wrap(~mode, scales = "free_y") +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  labs(
    title = "Pandemic Ridership Trends (2020–2021)",
    x = "Month",
    y = "Average Daily Ridership"
  ) +
  theme_minimal()

post =
full_mta |> 
  filter(
    period == "Post-Pandemic"
  ) |> 
  ggplot(aes(x = date, y = avg_daily_ridership)) +
  geom_line(linewidth = 1, color = "#08519c") +
  facet_wrap(~mode, scales = "free_y") +
  labs(
    title = "Post-Pandemic Ridership Trends (2022–2025)",
    x = "Month",
    y = "Average Daily Ridership"
  ) +
  theme_minimal()

combined = pre / pandemic / post

combined
```

Analyses for pre-pandemic, pandemic, and post-pandemic were organized into a facet grid showcasing `Average Daily Ridership` across six different modes of public transportation. 

Pre-pandemic average daily ridership remained expectedly variable.

Pandemic average daily ridership begins with a sharp decrease in ridership at the start of the pandemic and then gradually increases over time.

Post-pandemic average daily ridership patterns (for all modes of transit except `Bus`) also tend to gradually increase up until 2025. `Bus` ridership remains relatively variable from 2022 to 2025.

## Plotting Weekday vs. Weekend Ridership Volume during Pandemic vs Post-Pandemic
```{r}
pre_wknd =
daily_mta_clean |> 
  group_by(mode, weekend) |>
  ungroup() |> 
  filter(
    year %in% c(2020,2021)
  ) |> 
ggplot(aes(x = weekend, y = count, fill = weekend)) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.12, outlier.alpha = 0.2) +
  facet_wrap(~mode, scales = "free_y") +
  scale_fill_manual(values = c("Weekday" = "#E76F51", "Weekend" = "#2A9D8F")) +
  labs(
    title = "Weekday vs Weekend Ridership During Pandemic (2020–2021)",
    x = "",
    y = "Daily Ridership"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12)
  )

post_wknd =
daily_mta_clean |> 
  group_by(mode, weekend) |>
  ungroup() |> 
  filter(
    year %in% c(2022,2023, 2024, 2025)
  ) |> 
ggplot(aes(x = weekend, y = count, fill = weekend)) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.12, outlier.alpha = 0.2) +
  facet_wrap(~mode, scales = "free_y") +
  scale_fill_manual(values = c("Weekday" = "#E76F51", "Weekend" = "#2A9D8F")) +
  labs(
    title = "Weekday vs Weekend Ridership During Pandemic (2020–2021)",
    x = "",
    y = "Daily Ridership"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12)
  )


combined_weekend = pre_wknd / post_wknd
combined_weekend
```

Using the clean daily ridership dataset, `daily_mta_clean`, a violin plot was created of daily ridership for each mode of transportation for `Weekday` and `Weekend`, comparing ridership during the pandemic (2020-2021) and following the pandemic (2022-2025). 

Overall, for both pandemic and post-pandemic analyses, ridership remains decreased during the weekends. However, during the pandemic, overall ridership suffered a decrease with average daily ridership for the subway, for example, sitting at half the average daily ridership post-pandemic.

Pre-pandemic ridership data could not be assess in this way since the pre-pandemic dataset only contained monthly ridership data; weekday/weekend could not be differentiated using the monthly data. Therefore, the analysis is limited to the daily ridership dataset, which includes pandemic and onward.

# Grouped Analyses 

# COVID and MTA Comparison 

#### MTA Ridership and Covid-19 Cases

```{r}
# Average daily MYA Ridership 
ridership_modes =
  full_mta |>
  mutate(
    mode_group = case_when(
      mode == "Bus"    ~ "Bus",
      mode == "Subway" ~ "Subway",
      TRUE             ~ "Other"
    )
  ) |>
  group_by(date, mode_group) |>
  summarise(
    avg_daily_ridership = sum(avg_daily_ridership, na.rm = TRUE),
    .groups = "drop"
  )

ridership_plot =
  ridership_modes |>
  plot_ly(
    x     = ~date,
    y     = ~avg_daily_ridership,
    color = ~mode_group,
    name  = ~mode_group,
    legendgroup = ~mode_group,
    type  = "scatter",
    mode  = "lines"
  ) |>
  layout(
    title = list(
      text = "NYC MTA Ridership and COVID-19 Cases",
      x    = 0.5
    ),
    yaxis = list(title = "Average Daily Ridership"),
    showlegend = TRUE
  )

#COVID Daily PLOT visulation- same as above but without viridis 

covid_daily_plot =
  clean_covid_df |>
  plot_ly(
    x = ~date,
    y = ~case_count,
    type = "scatter",
    mode = "lines",
    line = list(color = "firebrick", width = 0.7),
    showlegend = FALSE   
  ) |>
  layout(
    yaxis = list(
      title = "Daily COVID-19 Cases (log scale)",
      type  = "log"
    ),
    xaxis = list(title = "Date"),
    shapes = lapply(1:nrow(events), function(i) {
      list(
        type = "rect",
        xref = "x",
        x0 = events$start[i],
        x1 = events$end[i],
        yref = "paper",
        y0 = 0,
        y1 = 1,
        fillcolor = "rgba(255, 0, 0, 0.15)",
        line = list(width = 0),
        layer = "below"
      )
    }),
    annotations = lapply(1:nrow(events), function(i) {
      list(
        x = events$start[i] + (events$end[i] - events$start[i]) / 2,
        y = 1.02,
        xref = "x",
        yref = "paper",
        text = events$event[i],
        showarrow = FALSE,
        font = list(size = 12)
      )
    })
  )

final_figure =
  subplot(
    ridership_plot,
    covid_daily_plot,
    nrows   = 2,
    shareX  = TRUE,
    titleX  = TRUE,
    titleY  = TRUE,
    heights = c(0.55, 0.45)
  ) |>
  layout(
    height    = 550,
    hovermode = "x unified"
  )

final_figure
```

- Had to stack the graphs instead of merging becasue the scale is so different 
- created hover so you can look at peaks and dips across both graphs 
- had to create the covid graph instead of using the DF saved earlier beause had to get rid of viridis it was giving me issues 
- Interpretation: Taken together, the two panels suggests that COVID-19 surges correspond with noticeable interruptions in ridership recovery / The figure highlights both the immediate impact of COVID waves and the broader, lasting transformation of mobility patterns. 
- To examine these interruptions more directly, I'm gonna create more specific surge periods—quantifying how ridership levels shifted during each wave relative to pre-pandemic baselines and seeing which modes were most resilient across disruptions.

#### Pandemic Recovery
```{r}
recovery =
  full_mta |>
  mutate(
    mode_group = case_when(
      mode == "Bus"    ~ "Bus",
      mode == "Subway" ~ "Subway",
      TRUE             ~ "Other"
    )
  ) |>
  group_by(mode_group) |>
  mutate(
    baseline_2018 = mean(avg_daily_ridership[year == 2018], na.rm = TRUE),
    pct_of_2018   = avg_daily_ridership / baseline_2018
  ) |>
  ungroup() |>
  filter(date >= as.Date("2020-06-01"))


recovery |> 
  ggplot(aes(x = date, y = pct_of_2018, color = mode_group)) +
    geom_smooth(se = FALSE, linewidth = 1.4) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      limits = c(0.4, 1.1)
    ) +
  labs(
    title = "Ridership Recovery Relative to 2018 Baseline",
    x = "Month",
    y = "% of 2018 Ridership",
    color = "Mode"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```

- Moved this down from individual analyses to grouped 

#### Ridership During Key Periods Compared to 2018 
```{r}
monthly_modes =
  full_mta |>
  mutate(
    mode_group = case_when(
      mode == "Bus"    ~ "Bus",
      mode == "Subway" ~ "Subway",
      TRUE             ~ "Other"
    )
  ) |>
  group_by(date, mode_group) |>
  summarise(
    ridership = sum(avg_daily_ridership, na.rm = TRUE),
    .groups = "drop"
  )

monthly_covid =
  clean_covid_df |>
  mutate(month = as.Date(cut(date, "month"))) |>
  group_by(month) |>
  summarise(
    monthly_cases = sum(case_count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(date = month)

monthly_joined =
  monthly_modes |>
  left_join(monthly_covid, by = "date")

monthly_surge =
  monthly_joined |>
  mutate(
    surge_period = case_when(
      date >= as.Date("2020-03-01") & date <= as.Date("2020-06-01") ~ "Lockdown wave",
      date >= as.Date("2020-12-01") & date <= as.Date("2021-03-01") ~ "Vaccine-rollout wave",
      date >= as.Date("2021-12-01") & date <= as.Date("2022-02-01") ~ "Omicron surge",
      TRUE ~ "Other months"
    )
  )

monthly_surge_filtered =
  monthly_surge |> 
  filter(surge_period != "Other months") |> 
  mutate(
    surge_period = factor(surge_period,
                          levels = c("Lockdown wave",
                                     "Vaccine-rollout wave",
                                     "Omicron surge"))
  )

baseline_2018 =
  monthly_surge |>  
  filter(year(date) == 2018) |>
  group_by(mode_group) |>
  summarise(
    baseline_2018 = mean(ridership, na.rm = TRUE),
    .groups = "drop"
  )

surge_summary =
  monthly_surge_filtered |>
  group_by(mode_group, surge_period) |>
  summarise(
    mean_ridership = mean(ridership, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(baseline_2018, by = "mode_group") |>
  mutate(
    pct_of_2018 = mean_ridership / baseline_2018
  )

plotly_surge =
  surge_summary |>
  plot_ly(
    x = ~surge_period,
    y = ~pct_of_2018,
    color = ~mode_group,
    type = "bar",
    hovertemplate =
      "<b>%{x}</b><br>"    %+%
      "Mode: %{color}<br>" %+%
      "Ridership: %{y:.1%} of 2018<extra></extra>"
  ) |>
  layout(
    barmode = "group",
    title = list(
      text = "MTA Ridership at Key Phases Compared to 2018 Baseline",
      x = 0.5
    ),
    xaxis = list(title = "COVID Surge Period"),
    yaxis = list(
      title = "% of 2018 Ridership",
      tickformat = ".0%",
      range = c(0, max(surge_summary$pct_of_2018) * 1.15)
    ),
    legend = list(title = list(text = "Mode"))
  )

plotly_surge
```

- Summarizes how MTA ridership (bus, subway, and other modes) changed across each key COVID-19 periods, shown as a percentage of 2018 daily baseline levels.
- Across all modes, the lock down wave produced the most dramatic decline (stay at home orders )
- During the vaccine-rollout wave, ridership partially recovered, depending on mode. (vaccination campaigns began, restrictions eased, and more people returned to work — but overall transit use remained far below pre-pandemic levels) 
- Although Omicron caused a spike in case counts, the city did not reinstate shutdown measures. (Workplaces, schools, and businesses remained open; vaccination rates were high; and the population’s behavioral response was less dramatic) which prolly explains why so much of ridership continued to return 

# Weather and MTA 

#### Average Monthly Temperature and MTA Ridership

```{r}

monthly_weather =
  weather_df |>
  mutate(month = as.Date(cut(date, "month"))) |>
  group_by(month) |>
  summarise(
    avg_temp = mean(average_temp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(date = month)

monthly_weather_mta =
  monthly_modes |>
  left_join(monthly_weather, by = "date")

ggplot(monthly_weather_mta,
       aes(x = avg_temp, y = ridership, color = mode_group)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE, size = 1.2) +
  labs(
    title = "Relationship Between Temperature and MTA Ridership (Monthly)",
    x = "Average Monthly Temperature (°C)",
    y = "Ridership",
    color = "Transit Mode"
  ) +
  theme_minimal(base_size = 14)


```

- Seems like weather does not have that much of an impact 

#### Extreme weather conditions (% Change compared to normal ridership days)

```{r}
weather_clean =
  weather_df |>
  mutate(
    avg_temp_c = average_temp,
    tmax_c,
    prcp_mm = prcp/10,    
    snow_mm = snow
  ) |>
  select(date, avg_temp_c, tmax_c, prcp_mm, snow_mm)

# defining the limits for extreme weather 
weather_extreme =
  weather_clean |>
  mutate(
    extreme =
      tmax_c >= 32.2 |     
      prcp_mm >= 25.4 |    
      snow_mm >= 127      
  ) |>
  select(date, extreme)

ridership_weather =
  ridership_modes |>
  left_join(weather_extreme, by = "date") |>
  mutate(extreme = replace_na(extreme, FALSE))

#computing percent change 
extreme_summary =
  ridership_weather |>
  group_by(mode_group, extreme) |>
  summarise(
    mean_ridership = mean(avg_daily_ridership, na.rm = TRUE),
    .groups = "drop_last"
  ) |>
  summarise(
    pct_drop =
      (mean_ridership[extreme == TRUE] -
       mean_ridership[extreme == FALSE]) /
       mean_ridership[extreme == FALSE],
    .groups = "drop"
  )

#plot 
extreme_plot =
  ggplot(extreme_summary,
         aes(x = mode_group, y = pct_drop, fill = mode_group)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c(
    "Bus"    = "#1b9e77",
    "Other"  = "#d95f02",
    "Subway" = "#7570b3"
  )) +
  labs(
    title = "Percent Change in Ridership on Extreme Weather Days",
    x = "Mode",
    y = "% Change (Extreme vs Normal)"
  ) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold"))

extreme_plot
```

# MTA x WEATHER x COVID Analysis- Linear Regression 

```{r}
monthly_all =
  monthly_joined |>      
  # Replaced missing COVID values with 0 (pre-pandemic months)
  mutate(monthly_cases = ifelse(is.na(monthly_cases), 0, monthly_cases)) |>
  
  left_join(
    weather_df |>
      mutate(month = as.Date(cut(date, "month"))) |>
      group_by(month) |>
      summarise(avg_temp = mean(average_temp, na.rm = TRUE),
                .groups = "drop") |>
      rename(date = month),
    by = "date"
  )

#Linear Model 
model_results =
  monthly_all |>
  group_by(mode_group) |>
  do(tidy(lm(ridership ~ avg_temp + monthly_cases, data = .)))

model_clean =
  model_results |>
  mutate(
    term = recode(
      term,
      "(Intercept)" = "Intercept",
      "avg_temp" = "Average Temperature (°C)",
      "monthly_cases" = "Monthly COVID-19 Cases"
    ),
    estimate  = round(estimate, 1),
    std.error = round(std.error, 1),
    statistic = round(statistic, 2),
    p.value   = signif(p.value, 3)
  ) |>
  arrange(mode_group, term)

# Making table look nicer 
model_clean |>
  rename(
    Term = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t value` = statistic,
    `p value` = p.value
  ) |>
  arrange(mode_group, Term) |>
  kable(
    caption = "Linear Models: Effect of Temperature and COVID-19 Cases on Ridership",
    digits = 3
  )

```

- We estimated separate linear regression models for bus, subway, and “other” transit modes to examine whether monthly average temperature and monthly COVID-19 case counts predicted ridership.
- Across all three modes, temperature was not a significant predictor of ridership (Bus: p = 0.172; Other: p = 0.961; Subway: p = 0.220), indicating that month-to-month weather variation did not meaningfully influence transit demand.
- COVID-19 case counts were strongly and consistently associated with reduced ridership for all (p<0.05)
- the models suggest that COVID-19 rather than ordinary weather fluctuations—were the dominant driver of pandemic-era ridership drops
- Temp has large coefficients but large standard erros so heir t values are not significant 




