---
title: "Individual and Grouped Analyses"
output: html_document
date: "2025-11-28"
---

```{r setup, include=FALSE}
library(lubridate)
library(tidyverse)
library(plotly)
library(patchwork)
library(dplyr)
library(scales)
library(knitr)
library(broom)
knitr::opts_chunk$set(fig.width = 10, fig.height = 6)
```
# Individual Analyses 

# COVID-19 ANALYSIS

#### Reading in the covid datafile.
```{r}
covid_df =
read.csv("data/covid_19.csv")
```
#### Separating the date variable into month, day, and year + arranging
```{r}
clean_covid_df =
  covid_df |>
  janitor::clean_names() |>
  mutate(
    date = lubridate::mdy(date_of_interest),
    case_count = as.numeric(gsub(",", "", case_count)),
    case_count_7day_avg = as.numeric(gsub(",", "", case_count_7day_avg)),
    hospitalized_count = as.numeric(gsub(",", "", hospitalized_count)),
    death_count = as.numeric(gsub(",", "", death_count))
  ) |>
  arrange(date)
```
#### Marking events for lockdown, rollout of vaccine, and the surge of omicron
```{r}
events = tibble(
  event = c("Lockdown", "Vaccine rollout", "Omicron surge"),
  start = as.Date(c("2020-03-15", "2020-12-14", "2021-12-01")),
  end   = as.Date(c("2020-06-15", "2021-03-01", "2022-02-15"))
)

max_y <- max(clean_covid_df$case_count, na.rm = TRUE)
```
#### Plotting the daily COVID-19 cases by Year
```{r}
clean_covid_df |> 
  plot_ly(
    x = ~date,
    y = ~case_count,
    type = "scatter",
    mode = "markers+lines",
    marker = list(
      color = ~case_count,
      colorscale = "Viridis",
      size = 5,
      opacity = 0.8,
      colorbar = list(title = "Daily Cases")
    ),
    line = list(color = "gray60", width = 0.5)
  ) |>
  layout(
    title = "NYC Daily COVID-19 Cases (Log Scale)",
    xaxis = list(
      title = "Year",
      range = c(as.Date("2019-01-01"), as.Date("2025-12-31"))
    ),
    yaxis = list(
      title = "Daily COVID-19 Cases (log scale)",
      type = "log"
    ),
    shapes = lapply(1:nrow(events), function(i) {
      list(
        type = "rect",
        xref = "x",
        x0 = events$start[i],
        x1 = events$end[i],
        yref = "paper",
        y0 = 0,
        y1 = 1,
        fillcolor = "rgba(255, 0, 0, 0.15)",
        line = list(width = 0),
        layer = "below"
      )
    }),
    annotations = lapply(1:nrow(events), function(i) {
      list(
        x = events$start[i] + (events$end[i] - events$start[i]) / 2,
        y = 1.03,
        xref = "x",
        yref = "paper",
        text = events$event[i],
        showarrow = FALSE,
        font = list(size = 12)
      )
    })
  )
```

- I changed it to log scale so that we can see all the peaks instead of it being concentrated at the x axis 
- Interpretation: The log-scale plot shows NYC’s daily COVID-19 cases from 2020–2025, with darker colors indicating lower counts and bright yellow marking the highest case days. The largest spike appears during the Omicron surge, far exceeding the earlier lockdown and vaccine-rollout waves even after log transformation. After Omicron, case levels drop substantially and settle into smaller, seasonal winter increases (which is expected) through 2023–2025, indicating a sustained decline in large outbreaks.

#### Plot Weekly COVID-19 Cases by Year
```{r}
clean_covid_df |> 
  plot_ly(
    x = ~date,
    y = ~case_count_7day_avg,
    type = "scatter",
    mode = "markers+lines",
    marker = list(
      color = ~case_count_7day_avg,
      colorscale = "Viridis",
      colorbar = list(title = "Weekly Avg Cases"),
      size = 5,
      opacity = 0.6
    ),
    line = list(
      color = "gray50",
      width = 0.6
    ),
    name = "7-day Average"
  ) |>
  layout(
    title = "NYC Weekly COVID-19 Cases (7-day Average, Log Scale)",
    xaxis = list(
      range = c(as.Date("2019-01-01"), as.Date("2025-12-31")),
      tickvals = as.Date(c("2019-01-01","2020-01-01","2021-01-01",
                           "2022-01-01","2023-01-01","2024-01-01","2025-01-01")),
      ticktext = c("2019","2020","2021","2022","2023","2024","2025"),
      title = "Year"
    ),
    yaxis = list(
      title = "7-day Average COVID-19 Cases (log scale)",
      type = "log"
    ),
    shapes = lapply(1:nrow(events), function(i) {
      list(
        type = "rect",
        xref = "x",
        x0 = events$start[i],
        x1 = events$end[i],
        yref = "paper",
        y0 = 0,
        y1 = 1,
        fillcolor = "rgba(255, 0, 0, 0.15)",
        line = list(width = 0),
        layer = "below"
      )
    }),
    annotations = lapply(1:nrow(events), function(i) {
      list(
        x = events$start[i] + (events$end[i] - events$start[i]) / 2,
        y = 1.0,
        xref = "x",
        yref = "paper",
        text = events$event[i],
        showarrow = FALSE,
        font = list(size = 12)
      )
    })
  )
```

- After the initial lockdown period, the weekly average shows a sharp and sustained decline in cases, reflecting the immediate impact of reduced mobility and strict public health measures. A similar post–vaccine-rollout decline appears in early 2021, where cases fall noticeably, reflecting the early effects of widespread vaccination and reduced transmission. 
- This again aligns with what we saw in the previous graph 

#### Correlation Heat Map of Cases, Hospitalization, and Deaths
```{r}
corr_matrix =
  clean_covid_df |>
    select(
      Cases = case_count,
      Hospitalizations = hospitalized_count,
      Deaths = death_count
    ) |>
    cor(use = "complete.obs")

plot_ly(
  x = colnames(corr_matrix),
  y = colnames(corr_matrix),
  z = corr_matrix,
  type = "heatmap",
  colorscale = "YlOrRd_r",
  zmin = min(corr_matrix),
  zmax = max(corr_matrix),
  text = round(corr_matrix, 2),
  texttemplate = "%{text}",
  textfont = list(color = "black", size = 16),
  showscale = TRUE
) |>
  layout(
    title = "Correlation Between NYC COVID-19 Cases, Hospitalizations, and Deaths",
    xaxis = list(title = ""),
    yaxis = list(title = "")
  )
```

- Interpretation: The correlation matrix shows that hospitalizations and deaths are strongly related, which reflects the expected clinical progression from severe illness to mortality. Cases have only a moderate correlation with hospitalizations and a weak correlation with deaths, likely because widespread vaccination, changing testing behavior, and the milder Omicron waves reduced the proportion of cases that became severe. 

# WEATHER ANALYSIS

#### Reading in weather data

```{r}
nyc_weather =
read.csv("data/nyc_weather.csv")
```

#### Cleaning up weather data

```{r}
weather_df =
  nyc_weather |> 
  janitor::clean_names() |> 
  mutate(
    date = as.Date(date),
    year = lubridate::year(date),
    month = lubridate::month(date),
    month = case_match(month,
                       1 ~ "January",
                       2 ~ "February",
                       3 ~ "March", 
                       4 ~ "April",
                       5 ~ "May",
                       6 ~ "June",
                       7 ~ "July",
                       8 ~ "August",
                       9 ~ "September", 
                       10 ~ "October",
                       11 ~ "November",
                       12 ~ "December"
    ),
    tmin_c = tmin / 10,
    tmax_c = tmax / 10,
    average_temp = (tmin_c + tmax_c) / 2,
  ) |> 
  filter(year %in% c(2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025))
```

- this data set stores temperature as tenths of degrees celcius so we had to fix that (ie 250 --> actually 25C) // Used central park as proxy 

#### Plot Average Temperature by Year

```{r}
weather_df |> 
  plot_ly(
    x = ~date,
    y = ~average_temp,
    type = "scatter",
    mode = "markers",
    marker = list(
      color = ~average_temp,
      colorscale = "Viridis",
      colorbar = list(title = "Temperature (°C)"),
      size = 5,
      opacity = 0.7
    )
) |>
  layout(
    title = "NYC Average Temperature by Year (2018-2025)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average Temperature (°C)")
  )
```

- The plot shows a clear and consistent seasonal pattern in NYC temperatures, with cold winters (often below 0°C) and warm summers that regularly exceed 25–30°C. 

#### Plot Snow Distribution by Year
```{r}
weather_df |>
  mutate(
    month = month(date, label = TRUE, abbr = FALSE),
    month = factor(month, 
                   levels = c("December", "January", "February", "March"))
  ) |>
  filter(
    snow > 0,
    month %in% c("December", "January", "February", "March")
  ) |>
  group_by(year, month) |>
  summarise(total_snow = sum(snow), .groups = "drop") |> 
  ggplot(aes(x = month, y = total_snow, fill = month)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~year, ncol = 2) +
  labs(
    title = "NYC Monthly Snowfall (mm) by Year (2018-2025)",
    x = "Month",
    y = "Total Snowfall (mm)",
    fill = ""
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- NYC snowfall is charted only for December through March because these four winter months account for virtually all measurable snowfall, making the analysis more meaningful. 
- Across these winters, snowfall is highly variable: 2018 and 2021 stand out as the snowiest seasons // 2019, 2020, 2024, and 2025 show more moderate but still noticeable accumulation spread across one or two winter months, 2022 has relatively light snowfall, and 2023 is nearly snow-free.
- uneven pattern highlights how NYC winters have become increasingly inconsistent

#### Plot Rainfall Distribution
```{r}
weather_df |>
  mutate(
    month = month(date, label = TRUE, abbr = FALSE),
    month = factor(month, 
                   levels = c("December", "January", "February", "March")),
    prcp_mm = prcp / 10
  ) |>
  filter(
    month %in% c("December", "January", "February", "March")
  ) |>
  group_by(year, month) |>
  summarise(total_prcp = sum(prcp_mm), .groups = "drop") |> 
  ggplot(aes(x = month, y = total_prcp, fill = month)) +
  geom_col(alpha = 0.8) +
  facet_wrap(~year, ncol = 2) +
  labs(
    title = "NYC Monthly Precipitation (mm) by Year (2018-2025)",
    x = "Month",
    y = "Total Precipitation (mm)",
    fill = ""
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- most years had precip concentrated around December and March, while January and February vary more widely. Some winters—such as 2018, 2019, and 2022—show consistently high precipitation across all four months, whereas others like 2020 and 2025 appear drier overall.

# MTA ANALYSIS

#### Read in MTA data
```{r}
daily_mta = 
  read.csv("data/mta.csv")

monthly_mta =
  read.csv("data/monthly_mta.csv")
```

#### Clean monthly and daily MTA datasets & join
```{r}
monthly_pre =
  monthly_mta |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(month),
    year = year(date),
    month_num = month(date),
    month_name = month(date, label = TRUE, abbr = FALSE),
    ridership = parse_number(ridership),

    mode = case_when(
      agency %in% c("NYCT Bus", "MTA Bus", "Bus") ~ "Bus",
      agency %in% c("Subway","MNR","LIRR","AAR","SIR") ~ agency,
      TRUE ~ NA_character_
    ),

    days_in_month = days_in_month(date),
    avg_daily_ridership = ridership / days_in_month
  ) |>
  filter(
    year >= 2018,
    year <= 2019,
    !is.na(mode)
  ) |>
  select(year, month_num, month_name, mode, avg_daily_ridership)

daily_mta_clean =
  daily_mta |>
  janitor::clean_names() |>
  mutate(
    date = mdy(date),
    year = year(date),
    month_num = month(date),
    month_name = month(date, label = TRUE, abbr = FALSE),
    count = parse_number(count),

    mode = case_when(
      mode %in% c("NYCT Bus", "MTA Bus", "Bus") ~ "Bus",
      mode %in% c("AAR","Subway","LIRR","MNR","SIR") ~ mode,
      TRUE ~ NA_character_
    )
  ) |>
  filter(year >= 2020, !is.na(mode)) |> 
  mutate(
    weekday = wday(date, label = TRUE),
    weekend = ifelse(weekday %in% c("Sat", "Sun"), "Weekend", "Weekday")
  )

monthly_post =
  daily_mta_clean |>
  group_by(year, month_num, month_name, mode) |>
  summarise(
    avg_daily_ridership = mean(count, na.rm = TRUE),
    .groups = "drop"
  )

full_mta =
  bind_rows(monthly_pre, monthly_post) |>
  arrange(year, month_num, mode) |>
  mutate(
    date = as.Date(paste(year, month_num, "01", sep = "-")),
    period = case_when(
      year < 2020 ~ "Pre-Pandemic",
      year == 2020 & month_num <= 2 ~ "Pre-Pandemic",
      year == 2020 & month_num >= 3 ~ "Pandemic",
      year == 2021 ~ "Pandemic",
      year >= 2022 ~ "Post-Pandemic"),
    period = factor(period, levels = c("Pre-Pandemic", "Pandemic", "Post-Pandemic"))
  )
```

#### Pre- vs. Post-Pandemic Modeling
```{r}
pre =
full_mta |> 
  filter(
    period == "Pre-Pandemic"
  ) |> 
  ggplot(aes(x = date, y = avg_daily_ridership)) +
  geom_line(linewidth = 1) +
  facet_wrap(~mode, scales = "free_y") +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  labs(
    title = "Pre-Pandemic Ridership Trends (2018–Feb 2020)",
    x = "Month",
    y = "Average Daily Ridership"
  ) +
  theme_minimal()

pandemic =
full_mta |> 
  filter(
    period == "Pandemic"
  ) |> 
ggplot(aes(x = date, y = avg_daily_ridership)) +
  geom_line(linewidth = 1, color = "red") +
  facet_wrap(~mode, scales = "free_y") +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y"
  ) +
  labs(
    title = "Pandemic Ridership Trends (2020–2021)",
    x = "Month",
    y = "Average Daily Ridership"
  ) +
  theme_minimal()

post =
full_mta |> 
  filter(
    period == "Post-Pandemic"
  ) |> 
  ggplot(aes(x = date, y = avg_daily_ridership)) +
  geom_line(linewidth = 1, color = "#08519c") +
  facet_wrap(~mode, scales = "free_y") +
  labs(
    title = "Post-Pandemic Ridership Trends (2022–2025)",
    x = "Month",
    y = "Average Daily Ridership"
  ) +
  theme_minimal()

combined = pre / pandemic / post

combined
```
* combined = all three graphs combined to show pre vs. pandemic vs. post

- Pre pandemic: modes were 

#### Pandemic Recovery
```{r}
recovery =
  full_mta |>
  mutate(
    mode_group = case_when(
      mode == "Bus"    ~ "Bus",
      mode == "Subway" ~ "Subway",
      TRUE             ~ "Other"
    )
  ) |>
  group_by(mode_group) |>
  mutate(
    baseline_2018 = mean(avg_daily_ridership[year == 2018], na.rm = TRUE),
    pct_of_2018   = avg_daily_ridership / baseline_2018
  ) |>
  ungroup() |>
  filter(date >= as.Date("2020-06-01"))


recovery |> 
  ggplot(aes(x = date, y = pct_of_2018, color = mode_group)) +
    geom_smooth(se = FALSE, linewidth = 1.4) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      limits = c(0.4, 1.1)
    ) +
  labs(
    title = "Ridership Recovery Relative to 2018 Baseline",
    x = "Month",
    y = "% of 2018 Ridership",
    color = "Mode"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```

- Maybe move this down to comparative analyses?

## Plotting Weekday vs. Weekend Ridership Volume during Pandemic vs Post-Pandemic
```{r}
pre_wknd =
daily_mta_clean |> 
  group_by(mode, weekend) |>
  ungroup() |> 
  filter(
    year %in% c(2020,2021)
  ) |> 
ggplot(aes(x = weekend, y = count, fill = weekend)) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.12, outlier.alpha = 0.2) +
  facet_wrap(~mode, scales = "free_y") +
  scale_fill_manual(values = c("Weekday" = "#E76F51", "Weekend" = "#2A9D8F")) +
  labs(
    title = "Weekday vs Weekend Ridership During Pandemic (2020–2021)",
    x = "",
    y = "Daily Ridership"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12)
  )

post_wknd =
daily_mta_clean |> 
  group_by(mode, weekend) |>
  ungroup() |> 
  filter(
    year %in% c(2022,2023, 2024, 2025)
  ) |> 
ggplot(aes(x = weekend, y = count, fill = weekend)) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.12, outlier.alpha = 0.2) +
  facet_wrap(~mode, scales = "free_y") +
  scale_fill_manual(values = c("Weekday" = "#E76F51", "Weekend" = "#2A9D8F")) +
  labs(
    title = "Weekday vs Weekend Ridership During Pandemic (2020–2021)",
    x = "",
    y = "Daily Ridership"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12)
  )


combined_weekend = pre_wknd / post_wknd
combined_weekend
```
- Limitation: Since our pre-pandemic dataset was `monthly_mta.csv` (monthly ridership data), we are unable to pull out specific weekday/weekend. Therefore, our analysis is limited to our daily `mta.csv` dataset, which includes pandemic and onward. 

# Grouped Analyses 

# COVID and MTA Comparison 

#### MTA Ridership and Covid-19 Cases

```{r}
# Average daily MYA Ridership 
ridership_modes =
  full_mta |>
  mutate(
    mode_group = case_when(
      mode == "Bus"    ~ "Bus",
      mode == "Subway" ~ "Subway",
      TRUE             ~ "Other"
    )
  ) |>
  group_by(date, mode_group) |>
  summarise(
    avg_daily_ridership = sum(avg_daily_ridership, na.rm = TRUE),
    .groups = "drop"
  )

ridership_plot =
  ridership_modes |>
  plot_ly(
    x     = ~date,
    y     = ~avg_daily_ridership,
    color = ~mode_group,
    name  = ~mode_group,
    legendgroup = ~mode_group,
    type  = "scatter",
    mode  = "lines"
  ) |>
  layout(
    title = list(
      text = "NYC MTA Ridership and COVID-19 Cases",
      x    = 0.5
    ),
    yaxis = list(title = "Average Daily Ridership"),
    showlegend = TRUE
  )

#COVID Daily PLOT visulation- same as above but without viridis 

covid_daily_plot =
  clean_covid_df |>
  plot_ly(
    x = ~date,
    y = ~case_count,
    type = "scatter",
    mode = "lines",
    line = list(color = "firebrick", width = 0.7),
    showlegend = FALSE   
  ) |>
  layout(
    yaxis = list(
      title = "Daily COVID-19 Cases (log scale)",
      type  = "log"
    ),
    xaxis = list(title = "Date"),
    shapes = lapply(1:nrow(events), function(i) {
      list(
        type = "rect",
        xref = "x",
        x0 = events$start[i],
        x1 = events$end[i],
        yref = "paper",
        y0 = 0,
        y1 = 1,
        fillcolor = "rgba(255, 0, 0, 0.15)",
        line = list(width = 0),
        layer = "below"
      )
    }),
    annotations = lapply(1:nrow(events), function(i) {
      list(
        x = events$start[i] + (events$end[i] - events$start[i]) / 2,
        y = 1.02,
        xref = "x",
        yref = "paper",
        text = events$event[i],
        showarrow = FALSE,
        font = list(size = 12)
      )
    })
  )

final_figure =
  subplot(
    ridership_plot,
    covid_daily_plot,
    nrows   = 2,
    shareX  = TRUE,
    titleX  = TRUE,
    titleY  = TRUE,
    heights = c(0.55, 0.45)
  ) |>
  layout(
    height    = 550,
    hovermode = "x unified"
  )

final_figure
```

- Had to stack the graphs instead of merging becasue the scale is so different 
- created hover so you can look at peaks and dips across both graphs 
- had to create the covid graph instead of using the DF saved earlier beause had to get rid of viridis it was giving me issues 
- Interpretation: Taken together, the two panels suggests that COVID-19 surges correspond with noticeable interruptions in ridership recovery / The figure highlights both the immediate impact of COVID waves and the broader, lasting transformation of mobility patterns. 
- To examine these interruptions more directly, I'm gonna create more specific surge periods—quantifying how ridership levels shifted during each wave relative to pre-pandemic baselines and seeing which modes were most resilient across disruptions.

#### Ridership During Key Periods Compared to 2018 
```{r}
monthly_modes =
  full_mta |>
  mutate(
    mode_group = case_when(
      mode == "Bus"    ~ "Bus",
      mode == "Subway" ~ "Subway",
      TRUE             ~ "Other"
    )
  ) |>
  group_by(date, mode_group) |>
  summarise(
    ridership = sum(avg_daily_ridership, na.rm = TRUE),
    .groups = "drop"
  )

monthly_covid =
  clean_covid_df |>
  mutate(month = as.Date(cut(date, "month"))) |>
  group_by(month) |>
  summarise(
    monthly_cases = sum(case_count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(date = month)

monthly_joined =
  monthly_modes |>
  left_join(monthly_covid, by = "date")

monthly_surge =
  monthly_joined |>
  mutate(
    surge_period = case_when(
      date >= as.Date("2020-03-01") & date <= as.Date("2020-06-01") ~ "Lockdown wave",
      date >= as.Date("2020-12-01") & date <= as.Date("2021-03-01") ~ "Vaccine-rollout wave",
      date >= as.Date("2021-12-01") & date <= as.Date("2022-02-01") ~ "Omicron surge",
      TRUE ~ "Other months"
    )
  )

monthly_surge_filtered =
  monthly_surge |> 
  filter(surge_period != "Other months") |> 
  mutate(
    surge_period = factor(surge_period,
                          levels = c("Lockdown wave",
                                     "Vaccine-rollout wave",
                                     "Omicron surge"))
  )

baseline_2018 =
  monthly_surge |>  
  filter(year(date) == 2018) |>
  group_by(mode_group) |>
  summarise(
    baseline_2018 = mean(ridership, na.rm = TRUE),
    .groups = "drop"
  )

surge_summary =
  monthly_surge_filtered |>
  group_by(mode_group, surge_period) |>
  summarise(
    mean_ridership = mean(ridership, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(baseline_2018, by = "mode_group") |>
  mutate(
    pct_of_2018 = mean_ridership / baseline_2018
  )

plotly_surge =
  surge_summary |>
  plot_ly(
    x = ~surge_period,
    y = ~pct_of_2018,
    color = ~mode_group,
    type = "bar",
    hovertemplate =
      "<b>%{x}</b><br>"    %+%
      "Mode: %{color}<br>" %+%
      "Ridership: %{y:.1%} of 2018<extra></extra>"
  ) |>
  layout(
    barmode = "group",
    title = list(
      text = "MTA Ridership at Key Phases Compared to 2018 Baseline",
      x = 0.5
    ),
    xaxis = list(title = "COVID Surge Period"),
    yaxis = list(
      title = "% of 2018 Ridership",
      tickformat = ".0%",
      range = c(0, max(surge_summary$pct_of_2018) * 1.15)
    ),
    legend = list(title = list(text = "Mode"))
  )

plotly_surge
```




