---
title: "Descriptive and Comparative Analyses"
output: html_document
date: "2025-11-28"
---

```{r setup, include=FALSE}
library(lubridate)
library(tidyverse)
library(plotly)
library(patchwork)
library(dplyr)
library(scales)
library(knitr)
library(broom)
knitr::opts_chunk$set(fig.width = 10, fig.height = 6)
```

# COVID-19 ANALYSIS

#### Reading in the covid datafile.
```{r}
covid_df =
read.csv("data/covid_19.csv")
```
#### Separating the date variable into month, day, and year + arranging
```{r}
clean_covid_df =
  covid_df |>
  janitor::clean_names() |>
  mutate(
    date = lubridate::mdy(date_of_interest),
    case_count = as.numeric(gsub(",", "", case_count)),
    case_count_7day_avg = as.numeric(gsub(",", "", case_count_7day_avg)),
    hospitalized_count = as.numeric(gsub(",", "", hospitalized_count)),
    death_count = as.numeric(gsub(",", "", death_count))
  ) |>
  arrange(date)
```
#### Marking events for lockdown, rollout of vaccine, and the surge of omicron
```{r}
events = tibble(
  event = c("Lockdown", "Vaccine rollout", "Omicron surge"),
  start = as.Date(c("2020-03-15", "2020-12-14", "2021-12-01")),
  end   = as.Date(c("2020-06-15", "2021-03-01", "2022-02-15"))
)

max_y <- max(clean_covid_df$case_count, na.rm = TRUE)
```
#### Plotting the daily COVID-19 cases by Year
```{r}
clean_covid_df |> 
  plot_ly(
    x = ~date,
    y = ~case_count,
    type = "scatter",
    mode = "markers+lines",
    marker = list(
      color = ~case_count,
      colorscale = "Viridis",
      size = 5,
      opacity = 0.8,
      colorbar = list(title = "Daily Cases")
    ),
    line = list(color = "gray60", width = 0.5)
  ) |>
  layout(
    title = "NYC Daily COVID-19 Cases (Log Scale)",
    xaxis = list(
      title = "Year",
      range = c(as.Date("2019-01-01"), as.Date("2025-12-31"))
    ),
    yaxis = list(
      title = "Daily COVID-19 Cases (log scale)",
      type = "log"
    ),
    shapes = lapply(1:nrow(events), function(i) {
      list(
        type = "rect",
        xref = "x",
        x0 = events$start[i],
        x1 = events$end[i],
        yref = "paper",
        y0 = 0,
        y1 = 1,
        fillcolor = "rgba(255, 0, 0, 0.15)",
        line = list(width = 0),
        layer = "below"
      )
    }),
    annotations = lapply(1:nrow(events), function(i) {
      list(
        x = events$start[i] + (events$end[i] - events$start[i]) / 2,
        y = 1.03,
        xref = "x",
        yref = "paper",
        text = events$event[i],
        showarrow = FALSE,
        font = list(size = 12)
      )
    })
  )
```

- I changed it to log scale so that we can see all the peaks instead of it being concentrated at the x axis 
- Interpretation: The log-scale plot shows NYC’s daily COVID-19 cases from 2020–2025, with darker colors indicating lower counts and bright yellow marking the highest case days. The largest spike appears during the Omicron surge, far exceeding the earlier lockdown and vaccine-rollout waves even after log transformation. After Omicron, case levels drop substantially and settle into smaller, seasonal winter increases (which is expected) through 2023–2025, indicating a sustained decline in large outbreaks.

#### Plot Weekly COVID-19 Cases by Year
```{r}
clean_covid_df |> 
  plot_ly(
    x = ~date,
    y = ~case_count_7day_avg,
    type = "scatter",
    mode = "markers+lines",
    marker = list(
      color = ~case_count_7day_avg,
      colorscale = "Viridis",
      colorbar = list(title = "Weekly Avg Cases"),
      size = 5,
      opacity = 0.6
    ),
    line = list(
      color = "gray50",
      width = 0.6
    ),
    name = "7-day Average"
  ) |>
  layout(
    title = "NYC Weekly COVID-19 Cases (7-day Average, Log Scale)",
    xaxis = list(
      range = c(as.Date("2019-01-01"), as.Date("2025-12-31")),
      tickvals = as.Date(c("2019-01-01","2020-01-01","2021-01-01",
                           "2022-01-01","2023-01-01","2024-01-01","2025-01-01")),
      ticktext = c("2019","2020","2021","2022","2023","2024","2025"),
      title = "Year"
    ),
    yaxis = list(
      title = "7-day Average COVID-19 Cases (log scale)",
      type = "log"
    ),
    shapes = lapply(1:nrow(events), function(i) {
      list(
        type = "rect",
        xref = "x",
        x0 = events$start[i],
        x1 = events$end[i],
        yref = "paper",
        y0 = 0,
        y1 = 1,
        fillcolor = "rgba(255, 0, 0, 0.15)",
        line = list(width = 0),
        layer = "below"
      )
    }),
    annotations = lapply(1:nrow(events), function(i) {
      list(
        x = events$start[i] + (events$end[i] - events$start[i]) / 2,
        y = 0.92,
        xref = "x",
        yref = "paper",
        text = events$event[i],
        showarrow = FALSE,
        font = list(size = 12)
      )
    })
  )
```

- After the initial lockdown period, the weekly average shows a sharp and sustained decline in cases, reflecting the immediate impact of reduced mobility and strict public health measures. A similar post–vaccine-rollout decline appears in early 2021, where cases fall noticeably, reflecting the early effects of widespread vaccination and reduced transmission. 
- This again aligns with what we saw in the previous graph 

#### Correlation Heat Map of Cases, Hospitalization, and Deaths
```{r}
corr_matrix =
  clean_covid_df |>
    select(
      Cases = case_count,
      Hospitalizations = hospitalized_count,
      Deaths = death_count
    ) |>
    cor(use = "complete.obs")

plot_ly(
  x = colnames(corr_matrix),
  y = colnames(corr_matrix),
  z = corr_matrix,
  type = "heatmap",
  colorscale = "YlOrRd_r",
  zmin = min(corr_matrix),
  zmax = max(corr_matrix),
  text = round(corr_matrix, 2),
  texttemplate = "%{text}",
  textfont = list(color = "black", size = 16),
  showscale = TRUE
) |>
  layout(
    title = "Correlation Between NYC COVID-19 Cases, Hospitalizations, and Deaths",
    xaxis = list(title = ""),
    yaxis = list(title = "")
  )
```

- Interpretation: The correlation matrix shows that hospitalizations and deaths are strongly related, which reflects the expected clinical progression from severe illness to mortality. Cases have only a moderate correlation with hospitalizations and a weak correlation with deaths, likely because widespread vaccination, changing testing behavior, and the milder Omicron waves reduced the proportion of cases that became severe. 

# WEATHER ANALYSIS

#### Reading in weather data

```{r}
nyc_weather =
read.csv("data/nyc_weather.csv")
```

#### Cleaning up weather data

```{r}
weather_df =
  nyc_weather |> 
  janitor::clean_names() |> 
  mutate(
    date = as.Date(date),
    year = lubridate::year(date),
    month = lubridate::month(date),
    month = case_match(month,
                       1 ~ "January",
                       2 ~ "February",
                       3 ~ "March", 
                       4 ~ "April",
                       5 ~ "May",
                       6 ~ "June",
                       7 ~ "July",
                       8 ~ "August",
                       9 ~ "September", 
                       10 ~ "October",
                       11 ~ "November",
                       12 ~ "December"
    ),
    tmin_c = tmin / 10,
    tmax_c = tmax / 10,
    average_temp = (tmin_c + tmax_c) / 2,
  ) |> 
  filter(year %in% c(2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025))
```

- this data set stores temperature as tenths of degrees celcius so we had to fix that (ie 250 --> actually 25C) // Used central park as proxy 













