---
title: "Combined Data-Analysis"

author: "Identifying the potential relationships of MTA, Weather, and Covid-19 combined"

output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

## COVID-19 and MTA  

### MTA Ridership and Covid-19 Cases
This plotly depicts MTA Average Daily Ridership from 2018 to 2025 atop Daily COVID-19 Cases. The MTA Average Daily Ridership plot shows three different curves–subway, bus, and other. Remaining transit mode categories are grouped into “other” due to limited counts.

Taken together, the two panels suggest that COVID-19 surges correspond with noticeable interruptions in ridership recovery. The figure highlights both the immediate impact of COVID-19 waves and the broader, lasting transformation of transit behavior in NYC. For example, during the Omicron surge in late 2021 to early 2022, the is a corresponding dip in Average Daily Ridership across all three transit mode categories.

To examine ridership interruptions more directly, it is necessary to quantify how ridership levels shifted during each wave relative to pre-pandemic baselines and to see which transit modes were most resilient across interruptions.

```{r, warning=FALSE, message=FALSE}
library(lubridate)
library(tidyverse)
library(plotly)
library(janitor)
library(scales)
library(broom)
library(knitr)
knitr::opts_chunk$set(fig.width = 10, fig.height = 6)

covid_df =
read.csv("data/covid_19.csv")

clean_covid_df =
  covid_df |>
  janitor::clean_names() |>
  mutate(
    date = lubridate::mdy(date_of_interest),
    case_count = as.numeric(gsub(",", "", case_count)),
    case_count_7day_avg = as.numeric(gsub(",", "", case_count_7day_avg)),
    hospitalized_count = as.numeric(gsub(",", "", hospitalized_count)),
    death_count = as.numeric(gsub(",", "", death_count))
  ) |>
  arrange(date)

events = tibble(
  event = c("Lockdown", "Vaccine rollout", "Omicron surge"),
  start = as.Date(c("2020-03-15", "2020-12-14", "2021-12-01")),
  end   = as.Date(c("2020-06-15", "2021-03-01", "2022-02-15"))
)

max_y <- max(clean_covid_df$case_count, na.rm = TRUE)

nyc_weather =
read.csv("data/nyc_weather.csv")


weather_df =
  nyc_weather |> 
  janitor::clean_names() |> 
  mutate(
    date = as.Date(date),
    year = lubridate::year(date),
    month = lubridate::month(date),
    month = case_match(month,
                       1 ~ "January",
                       2 ~ "February",
                       3 ~ "March", 
                       4 ~ "April",
                       5 ~ "May",
                       6 ~ "June",
                       7 ~ "July",
                       8 ~ "August",
                       9 ~ "September", 
                       10 ~ "October",
                       11 ~ "November",
                       12 ~ "December"
    ),
    tmin_c = tmin / 10,
    tmax_c = tmax / 10,
    average_temp = (tmin_c + tmax_c) / 2,
  ) |> 
  filter(year %in% c(2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025))


daily_mta = 
  read.csv("data/mta.csv")

monthly_mta =
  read.csv("data/monthly_mta.csv")

monthly_pre =
  monthly_mta |>
  janitor::clean_names() |>
  mutate(
    date = as.Date(month),
    year = year(date),
    month_num = month(date),
    month_name = month(date, label = TRUE, abbr = FALSE),
    ridership = parse_number(ridership),

    mode = case_when(
      agency %in% c("NYCT Bus", "MTA Bus", "Bus") ~ "Bus",
      agency %in% c("Subway","MNR","LIRR","AAR","SIR") ~ agency,
      TRUE ~ NA_character_
    ),

    days_in_month = days_in_month(date),
    avg_daily_ridership = ridership / days_in_month
  ) |>
  filter(
    year >= 2018,
    year <= 2019,
    !is.na(mode)
  ) |>
  select(year, month_num, month_name, mode, avg_daily_ridership)

daily_mta_clean =
  daily_mta |>
  janitor::clean_names() |>
  mutate(
    date = mdy(date),
    year = year(date),
    month_num = month(date),
    month_name = month(date, label = TRUE, abbr = FALSE),
    count = parse_number(count),

    mode = case_when(
      mode %in% c("NYCT Bus", "MTA Bus", "Bus") ~ "Bus",
      mode %in% c("AAR","Subway","LIRR","MNR","SIR") ~ mode,
      TRUE ~ NA_character_
    )
  ) |>
  filter(year >= 2020, !is.na(mode)) |> 
  mutate(
    weekday = wday(date, label = TRUE),
    weekend = ifelse(weekday %in% c("Sat", "Sun"), "Weekend", "Weekday")
  )

monthly_post =
  daily_mta_clean |>
  group_by(year, month_num, month_name, mode) |>
  summarise(
    avg_daily_ridership = mean(count, na.rm = TRUE),
    .groups = "drop"
  )

full_mta =
  bind_rows(monthly_pre, monthly_post) |>
  arrange(year, month_num, mode) |>
  mutate(
    date = as.Date(paste(year, month_num, "01", sep = "-")),
    period = case_when(
      year < 2020 ~ "Pre-Pandemic",
      year == 2020 & month_num <= 2 ~ "Pre-Pandemic",
      year == 2020 & month_num >= 3 ~ "Pandemic",
      year == 2021 ~ "Pandemic",
      year >= 2022 ~ "Post-Pandemic"),
    period = factor(period, levels = c("Pre-Pandemic", "Pandemic", "Post-Pandemic"))
  )

```

```{r, warning=FALSE, message=FALSE}
# Average daily MTA Ridership 
ridership_modes =
  full_mta |>
  mutate(
    mode_group = case_when(
      mode == "Bus"    ~ "Bus",
      mode == "Subway" ~ "Subway",
      TRUE             ~ "Other"
    )
  ) |>
  group_by(date, mode_group) |>
  summarise(
    avg_daily_ridership = sum(avg_daily_ridership, na.rm = TRUE),
    .groups = "drop"
  )

ridership_plot =
  ridership_modes |>
  plot_ly(
    x     = ~date,
    y     = ~avg_daily_ridership,
    color = ~mode_group,
    name  = ~mode_group,
    legendgroup = ~mode_group,
    type  = "scatter",
    mode  = "lines"
  ) |>
  layout(
    title = list(
      text = "NYC MTA Ridership and COVID-19 Cases",
      x    = 0.5
    ),
    yaxis = list(title = "Average Daily Ridership"),
    showlegend = TRUE
  )

#COVID Daily PLOT visulation- same as above but without viridis 

covid_daily_plot =
  clean_covid_df |>
  plot_ly(
    x = ~date,
    y = ~case_count,
    type = "scatter",
    mode = "lines",
    line = list(color = "firebrick", width = 0.7),
    showlegend = FALSE   
  ) |>
  layout(
    yaxis = list(
      title = "Daily COVID-19 Cases (log scale)",
      type  = "log"
    ),
    xaxis = list(title = "Date"),
    shapes = lapply(1:nrow(events), function(i) {
      list(
        type = "rect",
        xref = "x",
        x0 = events$start[i],
        x1 = events$end[i],
        yref = "paper",
        y0 = 0,
        y1 = 1,
        fillcolor = "rgba(255, 0, 0, 0.15)",
        line = list(width = 0),
        layer = "below"
      )
    }),
    annotations = lapply(1:nrow(events), function(i) {
      list(
        x = events$start[i] + (events$end[i] - events$start[i]) / 2,
        y = 1.02,
        xref = "x",
        yref = "paper",
        text = events$event[i],
        showarrow = FALSE,
        font = list(size = 12)
      )
    })
  )

final_figure =
  subplot(
    ridership_plot,
    covid_daily_plot,
    nrows   = 2,
    shareX  = TRUE,
    titleX  = TRUE,
    titleY  = TRUE,
    heights = c(0.55, 0.45)
  ) |>
  layout(
    height    = 550,
    hovermode = "x unified"
  )

final_figure
```

#### Pandemic Recovery 
```{r}
recovery =
  full_mta |>
  mutate(
    mode_group = case_when(
      mode == "Bus"    ~ "Bus",
      mode == "Subway" ~ "Subway",
      TRUE             ~ "Other"
    )
  ) |>
  group_by(mode_group) |>
  mutate(
    baseline_2018 = mean(avg_daily_ridership[year == 2018], na.rm = TRUE),
    pct_of_2018   = avg_daily_ridership / baseline_2018
  ) |>
  ungroup() |>
  filter(date >= as.Date("2020-06-01"))


recovery |> 
  ggplot(aes(x = date, y = pct_of_2018, color = mode_group)) +
    geom_smooth(se = FALSE, linewidth = 1.4) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    scale_y_continuous(
      labels = scales::percent_format(accuracy = 1),
      limits = c(0.4, 1.1)
    ) +
  labs(
    title = "Ridership Recovery Relative to 2018 Baseline",
    x = "Month",
    y = "% of 2018 Ridership",
    color = "Mode"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```
Based on this figure, none of the transit modes have fully recovered to pre-pandemic ridership levels. Bus ridership recovered the fastest early on, but flattened around 2023. This pattern can likely be attributed to the fact that buses serve essential, local, and short-distance trips, and have “fresher” air circulation than subways, for example. For subways, ridership shows a slower, steadier recovery path and still appears to be recovering steadily as commuting patterns normalize. That said, subway ridership still remains meaningfully below pre-pandemic levels. 

Ridership in the Other category has recovered more rapidly in the post-pandemic period–likely reflecting return-to-office policies and patterns on a hybrid or fully in-person basis. Whereas individuals that use the subway or bus are more likely to live in NYC and have alternative ways of commuting to work (i.e. walking, biking), individuals who use “Other” forms of public transit are likely those using long-distance and commuter rails (i.e. LIRR, MNR) and are thus more reliant on these forms to commute to work. This explains the pattern of more rapidly growing ridership during the post-pandemic period.


### Ridership During Key Periods Compared to 2018
This figure summarizes how MTA ridership across bus, subway, and other transit modes shifted during each major COVID-19 wave, expressed as a percentage of 2018 baseline levels. Across all modes, the initial lockdown period produced the steepest and most immediate collapse in ridership, reflecting stay-at-home orders and widespread service avoidance. During the vaccine-rollout wave, ridership began to rebound, though the extent of recovery varied by mode and remained well below pre-pandemic levels. 

In contrast, the Omicron surge, despite extremely high case counts, did not trigger comparable ridership declines. With schools, workplaces, and businesses remaining open and population immunity much higher, behavioral responses were far less severe, allowing ridership to continue its gradual return. Together, these patterns highlight how the relationship between COVID-19 transmission and mobility weakened over time as public policy and collective behavior adapted.
```{r}

monthly_modes =
  full_mta |>
  mutate(
    mode_group = case_when(
      mode == "Bus"    ~ "Bus",
      mode == "Subway" ~ "Subway",
      TRUE             ~ "Other"
    )
  ) |>
  group_by(date, mode_group) |>
  summarise(
    ridership = sum(avg_daily_ridership, na.rm = TRUE),
    .groups = "drop"
  )

monthly_covid =
  clean_covid_df |>
  mutate(month = as.Date(cut(date, "month"))) |>
  group_by(month) |>
  summarise(
    monthly_cases = sum(case_count, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(date = month)

monthly_joined =
  monthly_modes |>
  left_join(monthly_covid, by = "date")

monthly_surge =
  monthly_joined |>
  mutate(
    surge_period = case_when(
      date >= as.Date("2020-03-01") & date <= as.Date("2020-06-01") ~ "Lockdown wave",
      date >= as.Date("2020-12-01") & date <= as.Date("2021-03-01") ~ "Vaccine-rollout wave",
      date >= as.Date("2021-12-01") & date <= as.Date("2022-02-01") ~ "Omicron surge",
      TRUE ~ "Other months"
    )
  )

monthly_surge_filtered =
  monthly_surge |> 
  filter(surge_period != "Other months") |> 
  mutate(
    surge_period = factor(surge_period,
                          levels = c("Lockdown wave",
                                     "Vaccine-rollout wave",
                                     "Omicron surge"))
  )

baseline_2018 =
  monthly_surge |>  
  filter(year(date) == 2018) |>
  group_by(mode_group) |>
  summarise(
    baseline_2018 = mean(ridership, na.rm = TRUE),
    .groups = "drop"
  )

surge_summary =
  monthly_surge_filtered |>
  group_by(mode_group, surge_period) |>
  summarise(
    mean_ridership = mean(ridership, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(baseline_2018, by = "mode_group") |>
  mutate(
    pct_of_2018 = mean_ridership / baseline_2018
  )

plotly_surge =
  surge_summary |>
  plot_ly(
    x = ~surge_period,
    y = ~pct_of_2018,
    color = ~mode_group,
    type = "bar",
    hovertemplate =
      "<b>%{x}</b><br>"    %+%
      "Mode: %{color}<br>" %+%
      "Ridership: %{y:.1%} of 2018<extra></extra>"
  ) |>
  layout(
    barmode = "group",
    title = list(
      text = "MTA Ridership During COVID Surges<br><sup>Relative to 2018 baseline</sup>",
      x = 0.5
    ),
    xaxis = list(title = "COVID Surge Period"),
    yaxis = list(
      title = "% of 2018 Ridership",
      tickformat = ".0%",
      range = c(0, max(surge_summary$pct_of_2018) * 1.15)
    ),
    legend = list(title = list(text = "Mode"))
  )

plotly_surge
```

## Weather and MTA 
### Average Monthly Temperature and MTA Ridership
The plot shows that monthly temperature has very little influence on MTA ridership. Across subway, bus, and other modes, ridership stays fairly stable at both low and high temperatures, and the trend lines are nearly flat. Even during very cold or warm months, ridership does not shift in a clear or consistent way. Overall, weather appears to have only a small effect compared with the larger disruptions seen during the pandemic.

```{r, warning=FALSE, message=FALSE}
monthly_weather =
  weather_df |>
  mutate(month = as.Date(cut(date, "month"))) |>
  group_by(month) |>
  summarise(
    avg_temp = mean(average_temp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(date = month)

monthly_weather_mta =
  monthly_modes |>
  left_join(monthly_weather, by = "date")

ggplot(monthly_weather_mta,
       aes(x = avg_temp, y = ridership, color = mode_group)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE, size = 1.2) +
  labs(
    title = "Relationship Between Temperature and MTA Ridership (Monthly)",
    x = "Average Monthly Temperature (°C)",
    y = "Ridership",
    color = "Transit Mode"
  ) +
  theme_minimal(base_size = 14)
```

### Extreme weather conditions (% Change compared to normal ridership days)

```{r}
weather_clean =
  weather_df |>
  mutate(
    avg_temp_c = average_temp,
    tmax_c,
    prcp_mm = prcp/10,    
    snow_mm = snow
  ) |>
  select(date, avg_temp_c, tmax_c, prcp_mm, snow_mm)

# defining the limits for extreme weather 
weather_extreme =
  weather_clean |>
  mutate(
    extreme =
      tmax_c >= 32.2 |     
      prcp_mm >= 25.4 |    
      snow_mm >= 127      
  ) |>
  select(date, extreme)

ridership_weather =
  ridership_modes |>
  left_join(weather_extreme, by = "date") |>
  mutate(extreme = replace_na(extreme, FALSE))

#computing percent change 
extreme_summary =
  ridership_weather |>
  group_by(mode_group, extreme) |>
  summarise(
    mean_ridership = mean(avg_daily_ridership, na.rm = TRUE),
    .groups = "drop_last"
  ) |>
  summarise(
    pct_drop =
      (mean_ridership[extreme == TRUE] -
       mean_ridership[extreme == FALSE]) /
       mean_ridership[extreme == FALSE],
    .groups = "drop"
  )

#plot 
extreme_plot =
  ggplot(extreme_summary,
         aes(x = mode_group, y = pct_drop, fill = mode_group)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c(
    "Bus"    = "#1b9e77",
    "Other"  = "#d95f02",
    "Subway" = "#7570b3"
  )) +
  labs(
    title = "Percent Change in Ridership on Extreme Weather Days",
    x = "Mode",
    y = "% Change (Extreme vs Normal)"
  ) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold"))

extreme_plot
```


```{r extreme_legend, results='asis', echo=FALSE}
cat('
<div class="extreme-legend">
  <strong>Extreme Weather Definition</strong>
  <ul>
    <li><span class="legend-color bus"></span> <b>Maximum temperature ≥</b> 32.2°C</li>
    <li><span class="legend-color other"></span> <b>Precipitation ≥</b> 25.4 mm </li>
    <li><span class="legend-color subway"></span> <b>Snow ≥</b>  127 mm</li>
  </ul>
</div>
')
```

## MTA, WEATHER, and COVID Analysis- Linear Regression
The linear regression models examined how monthly temperature and COVID-19 case counts affected ridership across bus, subway, and other transit modes. Temperature was not a significant predictor for any mode, showing that normal month-to-month weather variation did not meaningfully influence transit demand. In contrast, COVID-19 case counts had strong and consistent negative associations with ridership across all modes, indicating that pandemic surges were the primary drivers of declines in transit use. Although some temperature coefficients were large, their standard errors were also large, which resulted in non-significant t-values. Overall, the analysis shows that COVID-19 had a much greater impact on ridership than typical weather changes.

```{r}
monthly_all =
  monthly_joined |>      
  # Replaced missing COVID values with 0 (pre-pandemic months)
  mutate(monthly_cases = ifelse(is.na(monthly_cases), 0, monthly_cases)) |>
  
  left_join(
    weather_df |>
      mutate(month = as.Date(cut(date, "month"))) |>
      group_by(month) |>
      summarise(avg_temp = mean(average_temp, na.rm = TRUE),
                .groups = "drop") |>
      rename(date = month),
    by = "date"
  )

#Linear Model 
model_results =
  monthly_all |>
  group_by(mode_group) |>
  do(tidy(lm(ridership ~ avg_temp + monthly_cases, data = .)))

model_clean =
  model_results |>
  mutate(
    term = recode(
      term,
      "(Intercept)" = "Intercept",
      "avg_temp" = "Average Temperature (°C)",
      "monthly_cases" = "Monthly COVID-19 Cases"
    ),
    estimate  = round(estimate, 1),
    std.error = round(std.error, 1),
    statistic = round(statistic, 2),
    p.value   = signif(p.value, 3)
  ) |>
  arrange(mode_group, term)

# Making table look nicer 
model_clean |>
  rename(
    Term = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t value` = statistic,
    `p value` = p.value
  ) |>
  arrange(mode_group, Term) |>
  kable(
    caption = "Linear Models: Effect of Temperature and COVID-19 Cases on Ridership",
    digits = 3
  )
```

